Capon Beamforming Applied on Moving Objects - An Analysis of Local Lateral Shift-Invariance

### 2013-12-12 HRUIofMovingObjects/jrnl_article.tex
Moving objects: done some writing. Updated figures.

\begin{abstract}
%\boldmath
If an ultrasound imaging system provides a presentation of a moving object which is sensitive to local spatial shifts, the system is said to be spatially locally shift-variant. This can for instance happen if the axial or lateral sampling is insufficient. The Capon beamformer has been shown to provide increased lateral resolution in ultrasound images. Increased lateral resolution should demand denser lateral sampling. In previous literature on Capon beamforming for medical ultrasound imaging however only single frame scenarios have been simulated. Temporal behaviour and effects caused by the increased resolution and lack of oversampling have therefore been neglected.

@@

\IEEEPARstart{I}{n} the last decade several authors have applied the Capon beamformer (also known as the minimum variance beamformer) for medical ultrasound imaging \cite{Synnevag2007, Vignon2008, Viola}. The adaptive nature of the method has been shown to greatly improve the lateral resolution and edge sharpness in ultrasound images, but with a high computational cost. In recent work we introduced an implementation of the Capon beamformer capable of processing a stream of images at interactive frame rates \cite{Asen}. It is therefore an interesting question how a moving object appears after Capon beamforming has been applied. Earlier work on Capon beamforming for medical ultrasound imaging has investigated the method's capabilities in single frames only, and temporal behaviour and effects has therefore been neglected.

@@

As mentioned, Capon beamforming has been shown to provide improved lateral resolution and edge sharpness in ultrasound images \cite{Synnevag2007, Synnevag2009, Chen2011}. The method accomplishes this by adjusting the array weights based on element-to-element covariance. Hence, the method yields data dependent weights. Capon beamforming is however also known for its suboptimal performance if e.g. the sample size available for covariance estimation is limited \cite{Mestre2006}, if correlated signals are present \cite{Widrow1982}, and if the assumed steering vector does not match the signal propagation vector \cite{Wax1996, Wax1996a}. This has led to the development of several robust versions of the Capon beamformer with e.g. spatial smoothing \cite{Shan1985} or diagonal loading \cite{JianLi2003} of the covariance estimate, or by applying a steering vector uncertainty set \cite{Lorenz2005, Rubsamen2013}. Spatial smoothing and diagonal loading, together with temporal smoothing of the covariance estimate, have been applied in earlier work on Capon beamforming for medical ultrasound imaging \cite{Synnevag2009}. The first two trade resolution for robustness and temporal smoothing helps to preserve speckle statistics for high resolution parameters \cite{Synnevag2007a}. However, the issue of having a mismatch between the assumed steering vector and the signal propagation vector has not been addressed directly.

A steering vector uncertainty set helps to reduce signal-nulling which appears if the signal of interest is present during the covariance estimation process and if this estimate is later used with a steering vector that does not match the signal propagation vector exactly. This self-nulling is actually how the Capon beamformer obtains its super resolution capabilities, but the signal can also get lost in the process. In \cite{JianLi2003} it was shown that most methods based on steering vector uncertainty sets are equivalent to diagonal loading where the loading factor depends on the steering vector uncertainty. In that way they make up for a given uncertainty by making the Capon beam wider. Hence, the maximal achievable resolution is decreased if the initial steering vector spacing was too coarse. In medical ultrasound imaging this initial set of steering vectors is govern by the number of receive beams. In this paper we will make use of an additional set of steering vectors to reduce the self-nulling effect by means of over sampling. The maximal achievable resolution is then maintained without the presence of severe self-nulling.

In \cite{Cox1973}, Cox derives formulas for the resolution of the Capon beamformer, and presents in great detail how the amount of mismatch between the selected steering vector and the signal propagation vector impacts the beamformer output in simple situations. Cox states the following important fact about the Capon beamformer (the $\vec{k}_3$ processor): 
\begin{quote}
"\textit{... a $\vec{k}_3$ processor requires more closely spaced beams than a ... $\vec{k}_1$-processor in order to avoid serious signal suppression effects being introduced on signals arriving from directions between the beams}", 
\end{quote}
where $\vec{k}_1$ is the delay-and-sum (DAS) beamformer. In Fig.\,\ref{fig:das_capon_beams} we have plotted the two-way response for DAS and the Capon beamformer using (26) and (32) in \cite{Cox1973}. A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to two times the required beam spacing ($1/64$ radians) using narrow band excitation. Located in the farfield of the sector is an object moving parallel to the array surface. The plot shows the scanned output when the object is located at plus and minus half the required beam spacing for DAS (\ref{eq:resolution}). No spatially white noise is present. Hence, the interference-noise matrix ($\sigma_0^2\mat{Q}$ in \cite{Cox1973}) is the identity. The effect of typical subarray averaging and diagonal loading ($L=32$ and $d=1/100$) has also been added to the Capon beamformer output. A diagonal loading of $1/100$ corresponds to a signal-to-noise ratio (SNR) of 20 dB before array gain is applied. The figure shows that a DAS beamformer, with the required beam spacing, could yield a $1.9$ dB scalloping loss \cite{Harris1978} if the object is moving parallel to the array. However if the Capon beamformer is used with the same beam density the scalloping loss could be as large as $27$ dB. A $1.9$ dB scalloping loss for the Capon beamforming is obtained for this specific example when the beam density is increased with a factor of 25.

@@

The solution to the minimization problem in (\ref{eq:capon_optimization_criteria}) is
\begin{align}\label{eq:capon_weights}
\vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}} \in \mathbb{C}^L.
\end{align}
Note that one matrix has to be constructed and inverted for each data vector $\vec{x}[n]$ received by the system. This is indeed computational demanding, but following the innovation in GPU computing, and by applying a novel method in order to reduce the size of the matrix inversion problem \cite{Nilsen2009}, real time processing is now possible \cite{Asen}. From (\ref{eq:capon_weights}) it is clear that complex data can lead to complex weights. The weights can therefore have an asymmetric K-space response which basically means that the main lobe and zeros does not have to be symmetric around the origin. The beam can therefore be micro-steered by phase rotation.

@@

Equation (\ref{eq:resolution}) was used to calculate the object position or beam spacing in Fig.\,\ref{fig:das_capon_beams}. Therefore, if (\ref{eq:resolution}) is not fulfilled the scalloping loss could become visible in the image and the system will no longer be locally laterally shift-invariant. In an ultrasound image with 256 gray levels and 50 dB dynamic range, a 1 dB loss corresponds to 5 gray levels. This is approximately equal to the visibility threshold (Weber fraction of 2\%) for a grayscale image, and a loss larger than 1 dB could therefore end up being visible to the observer. In our further analysis a system would be regarded as locally laterally shift-invariant if the scalloping loss is less than 1 dB. However, what eventually matters is whether the loss is visible in the image or not. %However, depending on viewing light conditions and e.g. the selected gray level mapping, a larger loss might be acceptable.

In order to visualize the local lateral shift-invariance of a given imaging system, we will make use of lateral shift-variance plots (LSV-plots), introduced by Hergum \textit{et al.} \cite{Hergum2007}. A LSV-plot is constructed for a given imaging system by imaging a point scatterer moving laterally at constant speed. For each lateral position the root-mean-square (RMS) of the image data is calculated in range, and the resulting beam profiles are stacked in to an image. For a given scan sequence there are twice as many point positions than there are receive lines. The LSV-plots in this paper are displayed as a contour plot with contours on -1, -2, -3, -6, -12 and -24 dB. The maximum for each beam profile is also marked with a black dot, and the amplitude variation among these points is presented in a subplot next to the LSV-plot. The LSV-plots based on simulations has been up-interpolated to 128 samples using cubic interpolation before presentation.

To further quantify the shift-invariance we have calculated the mean absolute error (MAE) of the maximum point position in each beam profile versus a straight line ($f(x)=x$). We will refer to this measure as P-MAE. The MAE of the maximum points' amplitude versus the data maximum is also measured. We refer to this measure as A-MAE. The first measure quantifies the amount of geometric distortion \cite{Hergum2007}, and the latter quantifies the peak gain variation or in other words the scalloping loss. The maximum amplitude deviation, or maximum scalloping loss, along this point trace is also calculated. We present all these numbers together with the LSV-plot.

A given imaging system will then be locally laterally shift-invariant if the contours are diagonal and if the amplitude variation among the maximum points is less than 1 dB. Then there is minimal geometric distortion (low P-MAE) and the scalloping loss will hopefully not be visible for the observer (low A-MAE). %Since both effects can occur at the same time it is important to have a measure not effected by geometric distortion.

Figure \ref{fig:das_shiftinvariant} presents such a LSV-plot for an imaging system using the DAS beamformer with a beam spacing equal to (\ref{eq:resolution}). The dotted vertical lines indicates the position of transmit-receive beam pairs. The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} with a probe setup equal to the M5S probe by GE Vingmed Ultrasound, Horten, Norway. The probe was excited with a 2.5 MHz and 1.5 cycles long pulse with a $\text{f-number}=3$ and corresponding cosine apodization on receive. The transmit focus in azimuth and elevation was configured to be equal and the point scatterer was moved through this focus. The lateral sampling density has been doubled prior to detection through coherent interpolation. We see that the contours in Fig.\,\ref{fig:das_shiftinvariant} are approximately diagonal (low P-MAE) and that the amplitude variation among the maximum points is minimal (low A-MAE). We therefore conclude that the system is locally laterally shift-invariant. In this, and the following LSV-plots based on simulations, the scan sequence always has a beam at zero degrees and the point scatterer is located in the center of this beam at one time instance.

The scalloping loss in Fig.\,\ref{fig:das_shiftinvariant} is smaller than what was predicted by Fig.\,\ref{fig:das_capon_beams}. To understand why note that Fig.\,\ref{fig:das_capon_beams} shows the scanned response for a narrowband system while the imaging system in Fig.\,\ref{fig:das_shiftinvariant} uses a broadband pulse. The beam in Fig.\,\ref{fig:das_shiftinvariant} will therefore be more similar to a squared gaussian function than the squared $sinc$ functions depicted in Fig.\,\ref{fig:das_capon_beams}. The imaging system in Fig.\,\ref{fig:das_shiftinvariant} also uses an expanding cosine window on receive.

@@

\section{Local Lateral Shift-Invariance of the Capon Beamformer}\label{sec:capon_LLSI}
We will now investigate the local lateral shift-invariance of the Capon beamformer with parameters equal to the ones used in earlier work \cite{Synnevag2009, Asen}. In Fig.\,\ref{fig:capon1} the same point scatterer response as presented in Fig.\,\ref{fig:das_shiftinvariant} has been weighted with Capon weights calculated using (\ref{eq:capon_weights}) with $L = M/2 = 48$, $K=1$ and $d=1/100$. As mentioned, the beam at zero degrees is the only one where the point scatterer is located in the center of the beam. The signal cancellation caused by the difference between the signal propagation vector and the assumed steering vector is therefore minimal when the point scatterer has zero lateral offset. From this plot we observe that the signal cancellation is high for all the other offsets. As a consequence the scalloping loss is tremendous, and not far from the predicted value in Fig\,\ref{fig:das_capon_beams}. %The difference between the predicted and the simulated loss might be caused by several factors. The most important one being that Fig.\,\ref{fig:das_capon_beams} shows a single frequency whereas the RMS operator used in Fig.\,\ref{fig:capon1} average a broadband signal in range. For a slice through the center of the pulse the scalloping loss is below -30dB.

The diagonal factor $d$ and the effective adaptive aperture size $L$ can be used to control the mainlobe width of the Capon beamformer. In both extreme cases, $d\gg1$ and $L=1$, the Capon beamformer becomes either a triangle or uniform-weighted DAS beamformer respectively, and a beamspacing according to (\ref{eq:resolution}) would then be sufficient. In Fig.\,\ref{fig:capon2} we show how the scalloping loss is reduced when the amount of diagonal loading is increased.

@@

\section{Oversampling Methods}\label{sec:methods}
We will now discuss different ways of improving the local lateral shift-invariance when the Capon beamformer is applied. From the previous section it should be clear that we somehow need to increase the lateral sampling to make the Capon imaging system locally laterally shift-invariant. In this section we will investigate how this oversampling should be conducted, how large it needs to be and how the proposed methods will affect the system frame rate and the processing complexity.

@@

\subsection{Oversampling by Parallel Receive Beamforming}
Parallel receive beamforming (PRB) means that multiple narrow receive beams (rx-beams) are constructed from one broad tx-beam. Then, by reducing the number of tx-beams and relying on multiple rx-beams in order to get sufficient sampling on receive, it is possible to achieve high frame rates in e.g. 3D cardiac ultrasound imaging. However, PRB also introduces geometrical distortion which to some degree can be compensated for by using e.g. synthetic transmit beamforming (STB) \cite{Hergum2007, Denarie2011, Denarie2013}. In \cite{Rabinovich2013} it is shown that STB works with Capon beamforming. In \cite{Asen} we also used STB in order to increase the beam density of a cardiac ultrasound recording. In the following discussion we consider corrected PRB as being equal to oversampling on transmit when it comes to image quality. The LSV-plots for PRB are therefore not included.

@@

\subsection{Oversampling by Phase Rotation}
The Capon beamformer formula in (\ref{eq:capon_weights}) gives a direct opportunity for oversampling using phase rotation. The steering vector $\vec{a}$ has in previous work been set to $\vec{1}$ since the ultrasound data are pre-delayed. However, the steering vector can also, as in narrow band applications, be varied over a set of pre-defined vectors. In a narrow band setting the beam can make a sweep across the whole sector by using phase rotation. This is also possible in broadband applications as long as the phase rotation is less than one wave length (See coarse-fine beamforming in \cite{Thomenius}). It is therefore possible to make a sweep between a given beam and its two nearest neighbours using only phase rotation without introducing large errors. Phase rotation is actually what the Capon beamformer use internally to micro-steer away from bright point scatterers and to enhance edges.

For a uniform linear array located along the x-axis the steering vector $\vec{a}_\theta$ should be calculated as 
\begin{align}
\vec{a}_\theta = 
\begin{bmatrix}
e^{-j\frac{2\pi}{\lambda_c}x_0\sin(\theta)} \\
e^{-j\frac{2\pi}{\lambda_c}x_1\sin(\theta)} \\
\vdots \\ 
e^{-j\frac{2\pi}{\lambda_c}x_{L-1}\sin(\theta)}
\end{bmatrix}
\end{align}
where $x_i$ is the element position and $\theta$ is swept from $-\Delta/2$ to $\Delta/2$ where $\Delta$ is as defined in (\ref{eq:resolution}). Note that the steering vector is calculated for a $L$ element subarray centered around the origin, even though the full array has $M$ elements. It is important that this subarray has the same element pitch, $x_i - x_{i-1}$, as the full array. %The steering error introduced by this steering vector for a frequency band $B$ is less than $B/2 \times \Delta/2$.

The steering angles can be spread out in many different ways. If the number of steering angles is $N_\theta$ we have selected a uniform distribution of $\theta$  with $N_\theta/2$ angles on each side of the original steering direction if $N_\theta$ is even, and $(N_\theta-1)/2$ if $N_\theta$ is odd.

Since the phase rotation method operates on the same data as before, there will be no reduction in the acquisition frame rate by applying this method. In addition, since the same number of covariance matrices are constructed and inverted, there is no extra cost involved here as well. The only additional cost will be the multiplication of the extra steering vectors with the inverted covariance matrix. If the equation $\mat{\hat{R}}\vec{a} = \vec{b}$, as in \cite{Asen}, is solved using the Gauss-Jordan algorithm, we can solve the problem for all steering vectors at once by row reducing the matrix $[\mat{\hat{R}},  \vec{a}_{\theta_1}, \vec{a}_{\theta_2}, \cdots \vec{a}_{\theta_{N_\theta}}]$ to the following form $[\mat{I},  \vec{b}_{\theta_1}, \vec{b}_{\theta_2}, \cdots \vec{b}_{\theta_{N_\theta}}]$. This is still an $O(L^3)$ operation.

@@

\section{In-vitro phantom data}\label{sec:res}
In Fig.\,\ref{fig:LSV_invitro} we present the result of applying the Capon beamformer on \textit{in-vitro} data with and without oversampling by phase rotation. The data was acquired using a modified Vivid E9 high-end cardiac ultrasound scanner by GE Vingmed Ultrasound, Horten, Norway, equipped with a M5S transducer operating in harmonic mode with a receive frequency of 3.7 MHz. The transmit beam density was set to match (\ref{eq:resolution}) based on this frequency. Note that in harmonic imaging the generated transmit beam at the harmonic frequency is a weighted version of the beam resulting from transmitting at the same frequency \cite{Fedewa2004}. The lateral sampling is therefore better than what was used in simulations. A tissue mimicking phantom containing wire targets and anechoic cysts was imaged, and the data was acquired while the probe was moved across the phantom manually. Fig.\,\ref{fig:image_das} shows the first frame out of 14 in the acquired loop. %Movies of the whole loop processed with DAS and Capon beamforming with and without ten times oversampling by phase rotation can be found in the supplementary material \multimedia{Videos}.

In Fig\,\ref{fig:image_das} one point scatterer has been selected (white box), and then used to generate LSV-plots. \comment{Todo: Add lateral coherent interpolation as done for the simulations? As for now speckle is not fully resolved without it.} Figure \ref{fig:phantom_das} shows the resulting LSV-plot when using the DAS beamformer. The scalloping loss is less than 1 dB, and the system is therefore locally laterally shift-invariant. As a consequence point scatterers have a consistent appearance when subjected to motion \multimedia{Video DAS}.

Figure \ref{fig:phantom_capon_LSV} shows the same for the Capon beamformer without oversampling. In the frame where the point scatterer is at -6 degrees the receive beam direction is very close to the signal propagation vector. Hence, we get a high amplitude value at this location. In almost all the other frames the signal loss is massive (10 dB). The result is blinking point scatterers when subjected to motion \multimedia{Video Capon}.

>>>>

\begin{abstract}
%\boldmath
If an ultrasound imaging system provides a presentation of a moving object which is sensitive to local spatial shifts, the system is said to be spatially locally shift-variant. This can for instance happen if the axial or lateral sampling is insufficient. The Capon beamformer has been shown to provide increased lateral resolution in ultrasound images. Increased lateral resolution should demand denser lateral sampling. However, in previous literature on Capon beamforming for medical ultrasound imaging only single frame scenarios have been simulated. Temporal behaviour and effects caused by the increased resolution and lack of oversampling have therefore been neglected.

@@

\IEEEPARstart{I}{n} the last decade several authors have applied the Capon beamformer (also known as the minimum variance beamformer) for medical ultrasound imaging \cite{Synnevag2007, Vignon2008, Viola}. The method has been shown to improve the lateral resolution and edge sharpness in ultrasound images, but at an increased computational expense. In recent work we introduced an implementation of the Capon beamformer capable of processing ultrasound images at interactive frame rates \cite{Asen}. Now it is therefore an interesting question how a moving object appears after Capon beamforming has been applied. Earlier work on Capon beamforming for medical ultrasound imaging has investigated the method's capabilities in single frames only, and temporal behaviour and effects has therefore been neglected.

@@

As mentioned, Capon beamforming has been shown to provide improved lateral resolution and edge sharpness in ultrasound images \cite{Synnevag2007, Synnevag2009, Chen2011}. The method accomplishes this by adjusting the array weights based on element-to-element covariance. Hence, the method yields data dependent weights. Capon beamforming is however also known for its suboptimal performance if e.g. the sample size available for covariance estimation is limited \cite{Mestre2006}, if correlated signals are present \cite{Widrow1982}, and if the assumed steering vector does not match the signal propagation vector \cite{Wax1996, Wax1996a}. This has led to the development of several robust versions of the Capon beamformer with e.g. spatial smoothing \cite{Shan1985} or diagonal loading \cite{JianLi2003} of the covariance estimate, or by applying a steering vector uncertainty set \cite{Lorenz2005, Rubsamen2013}. Spatial smoothing and diagonal loading, together with temporal smoothing of the covariance estimate, have been applied in earlier work on Capon beamforming for medical ultrasound imaging \cite{Synnevag2009}. The first two trade resolution for robustness and temporal smoothing helps to preserve speckle statistics for high resolution parameters \cite{Synnevag2007a}. However, the issue of having a mismatch between the assumed steering vector and the signal propagation vector has not been addressed directly for medical ultrasound imaging.

A steering vector uncertainty set has been shown to reduce signal-nulling which appears if the signal of interest is present during the covariance estimation process and if this estimate is later used with a steering vector that does not match the signal propagation vector exactly. This self-nulling is actually how the Capon beamformer obtains its super resolution capabilities, but the signal can also get lost in the process. In \cite{JianLi2003} it was shown that most methods based on steering vector uncertainty sets are equivalent to diagonal loading where the loading factor depends on the steering vector uncertainty. In that way they make up for a given uncertainty by making the Capon beam wider. Hence, the maximal achievable resolution is decreased if the initial steering vector spacing was too coarse. In medical ultrasound imaging this initial set of steering vectors is govern by the number of receive beams. In this paper we make use of an additional set of steering vectors to reduce the self-nulling effect by means of over sampling. The maximal achievable resolution is then maintained without the presence of severe self-nulling.

In \cite{Cox1973}, Cox derives formulas for the resolution of the Capon beamformer, and presents in great detail how the amount of mismatch between the selected steering vector and the signal propagation vector impacts the beamformer output in simple situations. Cox states the following important fact about the Capon beamformer [the $\vec{k}_3$ processor]: 
\begin{quote}
"\textit{... a $\vec{k}_3$ processor requires more closely spaced beams than a ... $\vec{k}_1$-processor in order to avoid serious signal suppression effects being introduced on signals arriving from directions between the beams}", 
\end{quote}
where $\vec{k}_1$ is the delay-and-sum (DAS) beamformer. In Fig.\,\ref{fig:das_capon_beams} we have plotted the two-way response for DAS and the Capon beamformer using (26) and (32) in \cite{Cox1973}. A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to two times the required beam spacing ($1/64$ radians) using narrow band excitation. Located in the farfield of the sector is an object moving parallel to the array surface. The plot shows the scanned output when the object is located at plus and minus half the required beam spacing for DAS (\ref{eq:resolution}). No spatially white noise is present. Hence, the interference-noise matrix ($\sigma_0^2\mat{Q}$ in \cite{Cox1973}) is the identity. The effect of typical subarray averaging and diagonal loading ($L=32$ and $d=1/100$) has also been added to the Capon beamformer output. A diagonal loading of $1/100$ corresponds to a signal-to-noise ratio (SNR) of 20 dB before array gain is applied. The figure shows that a DAS beamformer, with the required beam spacing, could yield a $1.9$ dB scalloping loss \cite{Harris1978} if the object is moving parallel to the array. However if the Capon beamformer is used with the same beam density the scalloping loss could be as large as $27$ dB. A $1.9$ dB scalloping loss for the Capon beamforming is obtained for this specific example when the beam density is increased with a factor of 25.

@@

The solution to the minimization problem in (\ref{eq:capon_optimization_criteria}) is
\begin{align}\label{eq:capon_weights}
\vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}} \in \mathbb{C}^L.
\end{align}
Note that one matrix has to be constructed and inverted for each data vector $\vec{x}[n]$ received by the system. %This is indeed computational demanding, but following the innovation in GPU computing, and by applying a novel method in order to reduce the size of the matrix inversion problem \cite{Nilsen2009}, real time processing is now possible \cite{Asen}. 
From (\ref{eq:capon_weights}) it is also clear that complex data [e.g. in-phase and quadrature data] can lead to complex weights. The weights can therefore have an asymmetric K-space response with a shifted main lobe and zeros at arbitrary positions. %The beam can therefore be micro-steered by phase rotation.

@@

Equation (\ref{eq:resolution}) was used to calculate the object position or required beam spacing in Fig.\,\ref{fig:das_capon_beams}. Therefore, if (\ref{eq:resolution}) is not fulfilled the scalloping loss could become visible in the image and the system will no longer be locally laterally shift-invariant. In an ultrasound image with 256 gray levels and 50 dB dynamic range, a 1 dB loss corresponds to 5 gray levels. This is approximately equal to the visibility threshold (Weber fraction of 2\%) for a grayscale image, and a loss larger than 1 dB could therefore end up being visible to the observer. %However, what eventually matters is whether the loss is visible in the image or not.%In our further analysis a system would be regarded as locally laterally shift-invariant if the scalloping loss is less than 1 dB.  %However, depending on viewing light conditions and e.g. the selected gray level mapping, a larger loss might be acceptable.

In order to visualize the local lateral shift-invariance of a given imaging system, we will make use of lateral shift-variance plots (LSV-plots), introduced by Hergum \textit{et al.} \cite{Hergum2007}. A LSV-plot is constructed for a given imaging system by imaging a point scatterer moving laterally at constant speed. For each lateral position the root-mean-square (RMS) of the image data is calculated in range, and the resulting beam profiles are stacked in to an image. %For a given scan sequence there are twice as many point positions than there are receive lines. 
The LSV-plots in this paper are displayed as a contour plot with contours on -1, -2, -3, -6, -12 and -24 dB. The maximum for each beam profile is also marked with a black dot, and the amplitude variation among these points is presented in a subplot next to the LSV-plot. The LSV-plots based on simulations has been up-interpolated to 128 samples in each direction using cubic interpolation if the number of receive beams is less than 128.

To further quantify the shift-invariance we have calculated the mean absolute error (MAE) between the maximum point position in each beam profile and a straight line ($f(x)=x$). We will refer to this measure as P-MAE. The MAE of the maximum points' amplitude versus the data maximum is also measured. We refer to this measure as A-MAE. The first measure quantifies the amount of geometric distortion \cite{Hergum2007}, and the latter quantifies the peak gain variation or in other words the scalloping loss. The maximum amplitude deviation, or maximum scalloping loss, along this point trace is also calculated. We present all these numbers together with the LSV-plot.

A given imaging system will then be locally laterally shift-invariant if the contours are diagonal and if the amplitude variation among the maximum points is less than 1 dB. Then there is minimal geometric distortion (low P-MAE) and the scalloping loss will not be visible for the observer (low A-MAE). %Since both effects can occur at the same time it is important to have a measure not effected by geometric distortion.

Figure \ref{fig:das_shiftinvariant} presents such a LSV-plot for an imaging system using the DAS beamformer with a beam spacing equal to (\ref{eq:resolution}). The dotted vertical lines indicates the position of transmit-receive beam pairs. The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} with a probe setup equal to the M5S probe by GE Vingmed Ultrasound, Horten, Norway. The probe was excited with a 2.5 MHz and 1.5 cycles long pulse. The transmit focus in azimuth and elevation was configured to be equal and the point scatterer was moved through this focus. We see that the contours in Fig.\,\ref{fig:das_shiftinvariant} are approximately diagonal (low P-MAE) and that the amplitude variation among the maximum points is minimal (low A-MAE). We therefore conclude that the system is locally laterally shift-invariant. In this, and the following LSV-plots based on simulations, the scan sequence always has a beam at zero degrees and the point scatterer is located in the center of this beam at one time instance.

%The scalloping loss in Fig.\,\ref{fig:das_shiftinvariant} is smaller than what was predicted by Fig.\,\ref{fig:das_capon_beams}. To understand why note that Fig.\,\ref{fig:das_capon_beams} shows the scanned response for a narrowband system while the imaging system in Fig.\,\ref{fig:das_shiftinvariant} uses a broadband pulse. The beam in Fig.\,\ref{fig:das_shiftinvariant} will therefore be more similar to a squared gaussian function than the squared $sinc$ functions depicted in Fig.\,\ref{fig:das_capon_beams}. The imaging system in Fig.\,\ref{fig:das_shiftinvariant} also uses an expanding cosine window on receive.

@@

\section{Local Lateral Shift-Invariance of the Capon Beamformer}\label{sec:capon_LLSI}
We will now investigate the local lateral shift-invariance of the Capon beamformer with parameters equal to the ones used in earlier work \cite{Synnevag2009, Asen}. In Fig.\,\ref{fig:capon1} the same point scatterer response as presented in Fig.\,\ref{fig:das_shiftinvariant} has been weighted with Capon weights calculated using (\ref{eq:capon_weights}) with $L = M/2 = 48$, $K=1$ and $d=1/100$. As mentioned, the beam at zero degrees is the only one where the point scatterer is located in the center of the beam. The signal cancellation caused by the difference between the signal propagation vector and the assumed steering vector is therefore minimal when the point scatterer is located at this position. From this plot we observe that the signal cancellation is high for all the other offsets. As a consequence the scalloping loss is tremendous, and not far from the predicted value in Fig\,\ref{fig:das_capon_beams}. %The difference between the predicted and the simulated loss might be caused by several factors. The most important one being that Fig.\,\ref{fig:das_capon_beams} shows a single frequency whereas the RMS operator used in Fig.\,\ref{fig:capon1} average a broadband signal in range. For a slice through the center of the pulse the scalloping loss is below -30dB.

The diagonal factor $d$ and the effective adaptive aperture size $L$ can be used to control the mainlobe width of the Capon beamformer. In both extreme cases, $d\gg1$ and $L=1$, the Capon beamformer becomes either a triangle or uniform-weighted DAS beamformer respectively, and a beam spacing according to (\ref{eq:resolution}) would then be sufficient. In Fig.\,\ref{fig:capon2} we show how the scalloping loss is reduced when the amount of diagonal loading is increased.

@@

\section{Oversampling Methods}\label{sec:methods}
In this section we will discuss different methods aimed at improving the local lateral shift-invariance for the Capon beamformer. From the previous section it should be clear that we somehow need to increase the lateral sampling to make the Capon imaging system locally laterally shift-invariant. In this section we will investigate how this oversampling should be conducted, how large it needs to be and how the proposed methods will affect the system frame rate and the processing complexity.

@@

\subsection{Oversampling by Parallel Receive Beamforming}
Parallel receive beamforming (PRB) means that multiple narrow receive beams (rx-beams) are constructed from one broad tx-beam. Then, by reducing the number of tx-beams and relying on multiple rx-beams in order to get sufficient sampling on receive, it is possible to achieve high frame rates in e.g. 3D cardiac ultrasound imaging. However, PRB also introduces geometrical distortion which to some degree can be compensated for by using e.g. synthetic transmit beamforming (STB) \cite{Hergum2007, Denarie2011, Denarie2013}. In \cite{Rabinovich2013} it is shown that STB works well with Capon beamforming. In \cite{Asen} we also used STB in order to increase the beam density of a cardiac ultrasound recording. In the following discussion we consider corrected PRB as being equal to oversampling on transmit when it comes to image quality. The LSV-plots for PRB are therefore not included.

@@

\subsection{Oversampling by Phase Rotation}
The Capon beamformer formula in (\ref{eq:capon_weights}) gives a direct opportunity for oversampling using phase rotation. The steering vector $\vec{a}$ has in previous work been set to $\vec{1}$ since the ultrasound data are pre-delayed. However, the steering vector can also, as in narrow band applications, be varied over a set of pre-defined vectors. In a narrow band setting the beam can make a sweep across the whole imaging sector by using phase rotation. This is also possible in broadband applications as long as the phase rotation is less than one pulse length (See coarse-fine beamforming in \cite{Thomenius}). It is therefore possible to make a sweep between a given beam and its two nearest neighbours using phase rotation without introducing large errors. Phase rotation is actually what the Capon beamformer use internally to micro-steer away from bright point scatterers and to enhance edges.

For a uniform linear array located along the x-axis the steering vector $\vec{a}_\theta$ should be calculated as 
\begin{align}
\vec{a}_\theta = 
\begin{bmatrix}
e^{-j\frac{2\pi}{\lambda_c}x_0\sin(\theta)} \\
e^{-j\frac{2\pi}{\lambda_c}x_1\sin(\theta)} \\
\vdots \\ 
e^{-j\frac{2\pi}{\lambda_c}x_{L-1}\sin(\theta)}
\end{bmatrix}
\end{align}
where $x_i$ is the element position and $\theta$ is swept from $-\Delta/2$ to $\Delta/2$ where $\Delta$ is as defined in (\ref{eq:resolution}). Note that the steering vector must be calculated for a $L$ element symmetric subarray, even though the full array has $M$ elements. This subarray should have the same element pitch, $x_i - x_{i-1}$, as the full array. %The steering error introduced by this steering vector for a frequency band $B$ is less than $B/2 \times \Delta/2$.

The steering angles can be spread out in many different ways. If the number of steering angles is $N_\theta$ we have selected a uniform distribution of $\theta$  with $N_\theta/2$ angles on each side of the original steering direction when $N_\theta$ is even.%, and $(N_\theta-1)/2$ [including the original steering direction] if $N_\theta$ is odd.

Since the phase rotation method operates on the same channel data as in Fig.\,\ref{fig:capon1}, there will be no reduction in the acquisition frame rate by applying this method. In addition, since the same number of covariance matrices are constructed and inverted, there is no extra cost involved here as well. The only additional cost will be the multiplication of the extra steering vectors with the inverted covariance matrix. If the equation $\mat{\hat{R}}\vec{a} = \vec{b}$, as in \cite{Asen}, is solved using the Gauss-Jordan algorithm, we can solve the problem for all steering vectors at once by row reducing the matrix $[\mat{\hat{R}},  \vec{a}_{\theta_1}, \vec{a}_{\theta_2}, \cdots \vec{a}_{\theta_{N_\theta}}]$ to the following form $[\mat{I},  \vec{b}_{\theta_1}, \vec{b}_{\theta_2}, \cdots \vec{b}_{\theta_{N_\theta}}]$. This is still an $O(L^3)$ operation.

@@

\section{In-vitro phantom data}\label{sec:res}
In Fig.\,\ref{fig:LSV_invitro} we present the result of applying the Capon beamformer on \textit{in-vitro} data with and without oversampling by phase rotation. The data was acquired using a modified Vivid E9 high-end cardiac ultrasound scanner by GE Vingmed Ultrasound, Horten, Norway, equipped with a M5S transducer operating in harmonic mode with a receive frequency of 3.7 MHz. The transmit beam density was set to match (\ref{eq:resolution}) based on this frequency. Note that in harmonic imaging the generated transmit beam at the harmonic frequency will be wider and have lower sidelobes than the beam resulting from transmitting at the harmonic frequency \cite{Fedewa2004}. The lateral sampling is therefore increased compared with our simulations. A tissue mimicking phantom containing wire targets and anechoic cysts was imaged, and the data was acquired while the probe was moved across the phantom manually. Fig.\,\ref{fig:image_das} shows the first frame out of 14 in the acquired loop. %Movies of the whole loop processed with DAS and Capon beamforming with and without ten times oversampling by phase rotation can be found in the supplementary material \multimedia{Videos}.

In Fig\,\ref{fig:image_das} one point scatterer has been selected (white box), and then used to generate LSV-plots. Figure \ref{fig:phantom_das} shows the resulting LSV-plot when using the DAS beamformer. The scalloping loss is less than 1 dB, and the system is therefore locally laterally shift-invariant. As a consequence point scatterers have a consistent appearance when subjected to motion \multimedia{Video DAS}.

Figure \ref{fig:phantom_capon_LSV} shows the same point scatterer response processed with the Capon beamformer without oversampling. For the frame where the point scatterer is at -6 degrees the receive beam direction is very close to the signal propagation vector. Hence, we get a high amplitude value at this location. In almost all the other frames the signal loss is massive (10 dB). The result is blinking point scatterers when subjected to motion \multimedia{Video Capon}.
